Robust Adaptation to Multi-Scale Climate Variability

### 2019-01-08 report-earths-future/paper-2018-robust-adaptation-cyclical-risk.tex
Update again before sending to Dave for comments

We assess estimation bias and variance for three cases.
First, we consider an idealized scenario where only secular change is present in the system and \gls{lfv} is fully damped.
Next, we consider the ``pre-industrial'' case where there is no secular change but \gls{lfv} modulates climate risk in time.
Finally, we consider a more realistic (though still idealized) case with both \gls{lfv} and secular change.
These three cases are illustrated in \cref{fig:example-fit}, which shows a single synthetic streamflow sequence generated with $N=50$ and $M=100$.
Projections with each of the three fitting models are also shown.
In the remainder of this section we present a more systematic analysis of each of these three cases.
\begin{figure}
  \includegraphics[width=\textwidth]{Example-NINO3-M100-N50.pdf}
  \caption{
    An illustration of the estimation procedure.
    A single streamflow sequence with $N=50$ and $M=100$ is shown for each of the three cases (secular only, \gls{lfv} only, and secular plus \gls{lfv}) considered.
    The blue line shows the observed sequence.
    The gray shading indicates the 50\% and 95\% confidence intervals using each of the three fitting methods discussed.
    The horizontal black line indicates the flood threshold.
  }\label{fig:example-fit}
\end{figure}

@@

Evaluating and implementing investments for climate risk mitigation demands making projections of climate risk, which exhibits both \gls{lfv} and secular trends, over the $M$-year project life of the instrument.
The success of this prediction will depend on the identifiability of different signals from an $N$-year sample, the time scales of \gls{lfv} relative to the project life of the instrument, and the uncertainty associated with the estimation.
In this paper we took a synthetic data approach to explore the implications of varying $M$ and $N$ in stylized scenarios which represent important features of real-world hydroclimate systems.

The stationary, linear trend, and \gls{hmm} models considered for estimation present an idealized alternative to more complex, physically based models for future risk projection.
They were chosen to represent the ability of simple models to capture two features of interest: \gls{lfv} and secular trends.
The models were assessed on their performance over a complete future period of length $M$ years; as discussed in \cref{sec:introduction} the bias and variance computed together represent the likelihood for over- or under-design of the instrument in a given climate scenario (see \cref{fig:conceptual-bias-variance}).
If more complex methods, such as model chains based on \glspl{gcm} and hydrologic models, are used, the sample size $N$ used in our experiments may be straightforwardly interpreted as a surrogate for the total informational uncertainty in the analysis.

The climate adaptation decisions which this analysis can inform are typically framed as economic cost-benefit analyses which discount future cash flows at some annual rate.
The application of a positive discount rate, as is typical in the public and private sectors, emphasizes the importance of predicting near-term risk.
Even ignoring this, however, if the uncertainty associated with the performance of the instrument increases into the future (as shown in \cref{fig:conceptual-sketch}) then adaptation strategies with a short $M$ are preferred.
This is consistent with recent emphasis on the importance of making decisions that are robust to a wide range of possible future climates \citep[\ie,][]{Haasnoot:2013im, Poff:2015jn}.

For projects where $M$ is sufficiently short, cyclical climate variability is dominant over the project planning period \citep{Jain:2001hz,Hodgkins:2017hw}.
The long periods and high amplitudes of the oscillations shown in \cref{fig:observed-lfv} and discussed in \cref{sec:intro-lfv} highlight the importance of adapting not only to secular change but also, simultaneously, to \gls{lfv}.

For projects with longer $M$, it becomes necessary to consider the role of secular change.
As illustrated schematically in \cref{fig:conceptual-sketch}, uncertainties as to future population, economic development, renewable energy technology, and greenhouse gas mitigation collectively lead to very high uncertainty in projections of future climate risk.
As the physical mechanisms cascade from global (\eg, global mean surface temperature) to regional  \citep[\eg, storm track position][]{Barnes:2015gl} and local (\eg, annual-maximum streamflows) scales, uncertainties compound and increase \citep{Dittes:2017he}.
We find that when making risk projections that consider secular change, mis-specification of a trend's mathematical form yields estimates which are increasingly unreliable as $M$ increases.
This is particularly important given that in real-world settings the true form of the trend is not known and thus ``nonstationary'' projections which assume a particular type of trend are likely to perform poorly \citep{Montanari:2014hl, Serinaldi:2015bq}.

Although the simulations presented here are neatly divided into secular change, \gls{lfv} only, and \gls{lfv} plus secular change, real-world hydroclimate time series exhibit \gls{lfv} on many timescales and several sources of (not necessarily linear) secular change, as shown in \cref{fig:observed-lfv}.
Identifying and modeling these different physical signals requires a large amount of information, represented here by the length of the $N$-year historical record.
In places where long records are not available, however, information from nearby locations, reanalysis products, or \acrshort{gcm} runs may increase the amount of available information \citep{Merz:2008eh}, effectively increasing $N$.

>>>>

We assess estimation bias and variance for three cases.
First, we consider an idealized scenario where only secular change is present in the system and \gls{lfv} is fully damped.
Next, we consider the ``pre-industrial'' case where there is no secular change but \gls{lfv} modulates climate risk in time.
Finally, we consider a more realistic (though still idealized) case with both \gls{lfv} and secular change.
These three cases are illustrated in \cref{fig:example-fit}, which shows a single synthetic streamflow sequence generated with $N=50$ and $M=100$.
Projections with each of the three fitting models are also shown.
This figure highlights that even where projections of average streamflow are unbiased, if the spread is too large then projection of the threshold exceedance probability may be too large.

In the remainder of this section we present a more systematic analysis of each of these three cases.
\begin{figure}
  \includegraphics[width=\textwidth]{Example-NINO3-M100-N50.pdf}
  \caption{
    An illustration of the estimation procedure.
    A single streamflow sequence with $N=50$ and $M=100$ is shown for each of the three cases (secular only, \gls{lfv} only, and secular plus \gls{lfv}) considered.
    The blue line shows the observed sequence.
    The gray shading indicates the 50\% and 95\% confidence intervals using each of the three fitting methods discussed.
    The horizontal black line indicates the flood threshold.
  }\label{fig:example-fit}
\end{figure}

@@

Evaluating and implementing investments for climate risk mitigation involves making projections of climate risk, which exhibits both \gls{lfv} and secular trends, over the $M$-year project life of the instrument.
The success of this prediction will depend on the identifiability of different signals from an $N$-year sample, the time scales of \gls{lfv} relative to the project life of the instrument, and the uncertainty associated with the estimation.
In this paper we took a synthetic data approach to explore the implications of varying $M$ and $N$ in stylized scenarios which represent important features of real-world hydroclimate systems.

For projects where $M$ is sufficiently short, cyclical climate variability is dominant over the project planning period \citep{Jain:2001hz,Hodgkins:2017hw}.
The long periods and high amplitudes of the oscillations shown in \cref{fig:observed-lfv} and discussed in \cref{sec:intro-lfv} highlight the \emph{importance} of identifying and predicting this variability.
However, one's \emph{ability} to do so depends on having a model of sufficient complexity to represent the processes that cause \gls{lfv}, and the data to fit the model.
If sufficient information is not available, then simple models which represent fewer processes may be preferred.

For projects with longer $M$, it becomes necessary to consider the role of secular change.
As illustrated schematically in \cref{fig:conceptual-sketch}, uncertainties as to future population, economic development, renewable energy technology, and greenhouse gas mitigation collectively lead to very high uncertainty in projections of future climate risk.
As the physical mechanisms cascade from global (\eg, global mean surface temperature) to regional  \citep[\eg, storm track position][]{Barnes:2015gl} and local (\eg, annual-maximum streamflows) scales, uncertainties compound and increase \citep{Dittes:2017he}.\footnote{Let's add a citation here from the climate literature}
We show that mis-specification of a trend's mathematical form yields estimates which are increasingly unreliable as $M$ increases, an important caution for the field of ``nonstationary'' flood frequency analysis as noted by \citet{Montanari:2014hl} and \citet{Serinaldi:2015bq}.

The stationary, linear trend, and \gls{hmm} models considered for estimation present an idealized alternative to more complex, physically based models for future risk projection.
For example, flood frequency analysis may join observations across time and space \citep{Lima:2016kd,Merz:2008eh} or apply model chains based on \glspl{gcm} and hydrologic models \citep[see][]{Merz:2014gf}.
We suggest, then, that the sample size $N$ defined in our experiments may be straightforwardly interpreted as a measure of the total informational uncertainty in the analysis; as $N$ increases, informational uncertainty decreases.

Similarly, real-world climate adaptation plans will typically include multiple instruments which may be placed in different locations and times in a sequential fashion.
Even if the planning period of a portfolio is long, the individual instruments within the portfolio may have short planning periods.
Since \cref{sec:results} shows that the bias and variance of climate risk projections tend to increase with $M$, the total bias and variance associated with sequencing 20 consecutive $M=\SI{5}{year}$ projects will be less than that associated with making a single $M=\SI{100}{year}$ project.
This effect will be exacerbated by the fact that if the first $M=\SI{5}{year}$ project is based on estimates with informational uncertainty $N$, the second will have $N+5$, the third $N+10$, and so on.

The climate adaptation decisions which this analysis can inform are typically framed as economic cost-benefit analyses which discount future cash flows at some annual rate.
The application of a positive discount rate, as is typical in the public and private sectors, emphasizes the importance of predicting near-term risk.
Consequently, projects with long $M$ must overcome future discounting, the potential for large bias or variance, and that all estimates are made with informational uncertainty $N$ rather than a sequence of $N$, $N+\Delta N, \ldots$.\footnote{Should we create a new section called summary for these last two paragraphs?}

\section{Summary}

In this paper we considered how the temporal structure of the climate affects the potential for successful prediction over a finite $M$-year future period.
We began with three premises, or observations, about the nature of climate risk:
\begin{enumerate*}[label= (\roman*)]
  \item that planning decisions are made with finite horizons;
  \item that climate risk varies on many scales; and
  \item that the dominant processes that control this risk depend on the planning period itself.
\end{enumerate*}
Although the simulations presented here are neatly divided into secular change, \gls{lfv} only, and \gls{lfv} plus secular change, real-world hydroclimate time series exhibit \gls{lfv} on many timescales and several sources of (not necessarily linear) secular change, as shown in \cref{fig:observed-lfv}.
Identifying and modeling these different physical signals requires a large amount of information, represented here by the length of the $N$-year historical record.
In places where long records are not available, however, information from nearby locations, reanalysis products, or \acrshort{gcm} runs may increase the amount of available information \citep{Merz:2008eh}, effectively increasing $N$.
